---
title: "Exploratory Data Analysis"
output:
  pdf_document
---

Here, we explore the trends in wages across various features and summarize our findings. 

# Data Loading

```{r}
list.of.packages <- c("tidyverse", "maps", "ggplot2", "dplyr")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

library('tidyverse')
library(ggplot2)
library(dplyr)
library(maps)
library(viridis) # For color scales in maps

set.seed(123) # Setting a seed for reproducibility

# Loading the processed data
data <- read.csv("eda_data_clean.csv")

```

# Target Variable Histograms

Here, we look at histograms for INCTOT and Log(INCTOT)
 
```{r}

# Histogram for INCTOT
ggplot(data, aes(x = INCTOT)) +
geom_histogram(binwidth = 5000, fill = "blue", color = "black") +
labs(title = "Histogram of Total Personal Income", x = "Total Income", y = "Frequency")

# Histogram for Log(INCTOT)
ggplot(data, aes(x = log_INCTOT)) + # Assuming the log column is named as Log.INCTOT
geom_histogram(binwidth = 0.1, fill = "green", color = "black") +
labs(title = "Histogram of Log Transformed Total Personal Income", x = "Log of Total Income", y
= "Frequency")

```

We observe that by log transformation, we are able to shift the concentration from the left side of the original plot to a distribution which is more central and suitable for our analysis.



# Plotting AGE and Education against Total Income

```{r}

# Scatter Plot for Age vs INCTOT
ggplot(data, aes(x = AGE, y = INCTOT)) +
geom_point(alpha = 0.6, color = "red") +
labs(title = "Scatter Plot of Age vs Total Personal Income", x = "Age", y = "Total Personal
Income")

# Scatter Plot for EDUC vs INCTOT
ggplot(data, aes(x = EDUC, y = INCTOT)) +
geom_point(alpha = 0.6, color = "darkblue") +
labs(title = "Scatter Plot of Educational Attainment vs Total Personal Income", x = "Educational
Attainment", y = "Total Personal Income")

# Violin Plot for SEX vs INCTOT
ggplot(data, aes(x = as.factor(SEX), y = INCTOT, fill = as.factor(SEX))) +
geom_violin() +
labs(title = "Violin Plot of Sex vs Total Personal Income", x = "Sex", y = "Total Personal
Income")

```

We are able to observe trends in Age and Education, where in we see a rise in income as Age increases and then after a certain point (retirment), the income goes down. In case of Education, we see as the degree attained is better, the income is also generally better.

There are no stark differences for the gender plot. However, we can see that Males(1) have a slightly higher concentration on higher personal income as compared to Female(2).


# State-wise Trends

We plot the average wages across various states on the US map.


```{r}

us_map <- map_data("state")

# Aggregate data to get average INCTOT by STATEFIP
state_income <- data %>%
group_by(STATEFIP, STATE) %>%
summarise(Avg_INCTOT = mean(INCTOT, na.rm = TRUE))

state_income$STATE <- toupper(state_income$STATE)
us_map$region <- toupper(us_map$region)

map_data_merged <- merge(us_map, state_income, by.x = "region", by.y = "STATE", all.x = TRUE)
map_data_merged <- map_data_merged %>% select(-subregion)
map_data_merged_clean <- na.omit(map_data_merged)


# Plotting the map
ggplot(data = map_data_merged_clean) +
  geom_polygon(aes(x = long, y = lat, group = group, fill = Avg_INCTOT)) +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Average Wages") +
  labs(title = "Average Wages per State") +
  theme_void()

```

We observe the typical states where we would see large incomes, on eastern coast, New York and on the Western Coast, Seattle and California


# Top Metropolitans, Occupations, and Industry

Here, we aggregate the data to get the top 10 areas, occupations and indsutry based on average wages.


```{r}

# Top 10 Metropolitan Areas by Average Salary
data %>%
group_by(MET) %>%
summarise(Avg_Salary = mean(INCTOT, na.rm = TRUE)) %>%
arrange(desc(Avg_Salary)) %>%
head(10)

```


```{r}

data %>%
group_by(OCC_VAL) %>%
summarise(Avg_Salary = mean(INCTOT, na.rm = TRUE)) %>%
arrange(desc(Avg_Salary)) %>%
head(10)

data %>%
group_by(OCC_CATEG) %>%
summarise(Avg_Salary = mean(INCTOT, na.rm = TRUE)) %>%
arrange(desc(Avg_Salary)) %>%
head(10)

```




```{r}
data %>%
group_by(IND_VAL) %>%
summarise(Avg_Salary = mean(INCTOT, na.rm = TRUE)) %>%
arrange(desc(Avg_Salary)) %>%
head(10)

data %>%
group_by(IND_CATEG) %>%
summarise(Avg_Salary = mean(INCTOT, na.rm = TRUE)) %>%
arrange(desc(Avg_Salary)) %>%
head(10)

```

# Saving Data for modelling

Finally, once we have utilized the required columns for EDA, we save a data with only the required features for modelling

```{r}
data1 <- data %>% select(-METFIPS, -MET, -STATE, 
                         -OCC, -OCC_VAL, -OCC_CATEG, 
                         -IND, -IND_VAL, -IND_CATEG)

write.csv(data1, "model_data_clean.csv", row.names = FALSE)

```

